Likelihood-Based Inference for Generalized Linear Mixed Models: Inference with Package \pkg{glmm}

### 2017-02-07 JSS/article.tex
added english sentences, need to add code

\subsection{Formatting the data}
\subsection{Using the model-fitting function}
\subsection{Using the summary output}
\subsection{Performing additional inference}

>>>>

\subsection{Formatting the data}

The following vectors can be used to fit a generalized linear mixed model using the \texttt{glmm} package. These vectors can be contained in a data frame, but they do not need to be.
\begin{enumerate}
\item A response vector. If your response is Poisson, then the entries in the response vector must be natural numbers. If your response is Bernoulli, then the entries in the response vector must be $0$ and $1$. (For this version of \texttt{glmm}, these are the only two response types possible. If you need to fit a model with a different response, contact me.) 
\item At least one vector that will be used for defining the random effects' design matrix. For this version of \texttt{glmm}, the vector(s) should be class \texttt{factor}.
\item Vector(s) that will be used for defining the fixed effects' design matrix. The vector(s) can be of class \texttt{factor} or \texttt{numeric}. 
\end{enumerate}
The first two types of vectors described in the list are required. The last type is optional. That is, the minimum requirement to fit a \texttt{glmm} model is the response vector and one vector for defining the random effects' design matrix. \\

We use the \texttt{salamander} dataset as an example in this vignette. For your convenience, it is already included in the \texttt{glmm} package. The data arose from an experiment conducted at the University of Chicago in 1986 and were first presented by \citet[section 14.5]{mcc:nelder:1989}. Scientists paired female and male salamanders of two types (Rough Butt and White Side) and collected data on whether or not they mated.  \\

The variable \texttt{Mate} tells us whether the pair of salamanders mated. The value is $1$ if they successfully mated and $0$ if they did not. The variable \texttt{Cross} describes the type of female and male salamander. For example, \texttt{Cross = W/R} indicates a White Side female was crossed with a Rough Butt male. The variable \texttt{Female} contains the identification number of the female salamander, and the variable \texttt{Male} contains the identification number of the male salamander.\\

The first R  command shown below gives us access to the \texttt{glmm} package and and all of its commands. The second line of code gives us access to the \texttt{salamander} data frame.  The next three commands help us begin to understand the data. We have four variables: \texttt{Mate}, \texttt{Cross}, \texttt{Female}, and \texttt{Male}. The summary shows us \texttt{Mate} is numeric, \texttt{Cross} is a factor with four levels, \texttt{Female} is a factor, and \texttt{Male} is a factor.

\begin{CodeChunk}
\begin{CodeInput}
first input first line
first input second line
\end{CodeInput}
\begin{CodeOutput}
output of first input
\end{CodeOutput}
\begin{CodeInput}
second input
\end{CodeInput}
\end{CodeChunk}

\subsection{Using the model-fitting function}

  Following Model A from \citet{karim:zeger:1992}, we set \texttt{Mate} as the response, \texttt{Cross} as the fixed effect variable, and \texttt{Female} and \texttt{Male} as the random effect variables. That is, we would like to fit a generalized linear mixed model with a logit link (because the response is Bernoulli). We will have four fixed effect parameters ($\beta_{R/R}, \beta_{R/W},\beta_{W/R},\beta_{W/W}$). There is likely to be variability among the females and variability among the males. That is, some females will be more likely to mate than other females, and we would like the model to reflect the tendencies of the individual salamanders. We incorporate this into the model by including a random effect for each female salamander and a random effect for each male salamander. We believe the female salamanders' random effects are i.i.d. draws from $N(0, \nu_F)$, where $\nu_F$ is an unknown parameter to be estimated. Similarly, we believe the male salamanders' random effects are i.i.d. draws from $N(0,\nu_M)$, where $\nu_M$ is an unknown parameter to be estimated. Finally, we believe the female and male random effects are independent of one another. \\

In the following code, we  fit the model using the \texttt{glmm} command and save the model under the name  \texttt{sal}. Because \texttt{Mate} is our response, it is on the left of the $\sim$. We want to have a fixed effect for each of the four levels of \texttt{Cross}, so we type \texttt{Mate $\sim$ 0 + Cross}. Because \texttt{Cross} is a factor, typing \texttt{Mate $\sim$ Cross} would fit an equivalent model. \\

Next, the \texttt{random} list creates the design matrices for the random effects. Since we want two random effects for each cross (one from the female salamander and one from the male salamander), we type \texttt{list($\sim$ 0 + Female, $\sim$ 0 + Male)}. We  include the \texttt{0} because we want our random effects to be centered at 0. Almost always, you will want your random effects to have mean 0. \\

 Following the \texttt{random} list, the argument \texttt{varcomps.names} allows us to name the list of variance components. In the \texttt{random} list, we have placed the females first. Therefore, the order of the variance components names are first ``F'' and then ``M.''  \\

Next,  we specify the name of our data set. This is an optional argument. If the data set is not specified, \texttt{glmm} looks to the parent environment for the variables you have referenced. \\

After the name of the data set, we need to specify the type of the response. In the salamander mating example, the response is binary: the salamanders either mated or they did not. Therefore, the family is \texttt{bernoulli.glmm}. If your response is a count, then the family is \texttt{poisson.glmm}.\\

Next, we specify our Monte Carlo sample size \texttt{m}. The general rule is the larger the Monte Carlo sample size, the more accurate the Monte Carlo likelihood approximation (MCLA) will be, and the more accurate the resulting Monte Carlo maximum likelihood estimates (MCMLEs) will be. Ideally, you want the largest \texttt{m} that time allows. For this vignette, we have chosen a Monte Carlo sample size that allows for quick computation. If you are interested in accuracy in the resulting estimates for the salamander model, we suggest a larger Monte Carlo sample size.  \\

%Finally, we can decide whether we would like additional output (see details in section \ref{sec:otherstuff}). If we would like the additional output, we type \texttt{debug = TRUE}. The default is \texttt{debug = FALSE}.\\

We put this all together in the following commands. Note that we set the seed so that we can have reproducible results. In other words, if you set your seed to the same number and type the exact command listed below, your results should be identical to those listed here. Additionally, the \texttt{proc.time} commands have been used to give you an idea of how quickly the model can be fit. The times shown here are from fitting a model on an ultrabook that cost 500 USD in 2013.

\subsection{Using the summary output}
The \texttt{summary} command displays
\begin{itemize}
\item the function call (to remind you of the model you fit).
\item the link function.
\item the fixed effect estimates, their standard errors (calculated using observed Fisher information), their \texttt{z value} test statistics (testing whether the coefficients are significantly different from zero), the test's p-values, and the R-standard significance stars (optional).
\item the variance component estimates, their standard errors (calculated using observed Fisher information), their \texttt{z value} test statistics (testing whether the coefficients are significantly different from zero), the test's p-values, and the R-standard significance stars (optional).
\end{itemize}

Note that the p-value for the fixed effects is calculated using a two-sided alternative hypothesis ($H_A: \beta \neq 0$) while the p-value for the variance components is calculated using a one-sided alternative hypothesis ($H_A: \nu >0$) because variance components must be nonnegative.\\

R stuff here

Looking at our output, we can see that the type of cross significantly affects the salamanders' odds of mating. Additionally, both the variance components are significantly different from zero and should be retained in the model.\\

The summary provides the estimates needed to write our model. First, we establish a little notation. Let $\pi_i$ represent the probability of successful mating for salamander pair $i$. Let $I()$ be an indicator function, so that $I\text{(Cross=R/R)}$ is $1$ when the variable \texttt{Cross = R/R} and 0 otherwise. Let $u_i^F$ represent the random effect from the female salamander in the $i$th pair. Let $u_i^M$ represent the random effect from the male salamander in the $i$th pair. Since the response is Bernoulli, the canonical link is the log odds of successful mating. Using this notation, we write the model as follows.\\
\begin{align*}
\log \left( \dfrac{\pi_i}{1-\pi_i} \right) &= 0.956 * I\text{(Cross=R/R)} + 0.2805 * I\text{(Cross=R/W)}\\
 &+ \ensuremath{-1.8968} * I\text{(Cross=W/R)}
+ 0.9723 * I\text{(Cross=W/W)}\\
 &+ u_i^F+u_i^M\\
u_i^F \overset{i.i.d.}{\sim} &N(0,1.288)\\
u_i^M \overset{i.i.d.}{\sim} &N(0,1.084)\\
\end{align*}

Recall that \texttt{m}  in the above model was chosen for convenience to save time. The resulting parameter estimates have a little too much variability. If we increase \texttt{m}, the Monte Carlo standard error decreases.

\subsection{Calculating confidence intervals}

\subsection{Calculating Monte Carlo standard errors}

\subsection{Performing additional inference}
